BSTERGMs: Bayesian Separable-Temporal Exponential family Graph Models

### 2020-12-11 doc/paper_format.tex
directory adj

\ifcase 1  % choose 0=slides, 1=article, 2=refart
   \documentclass[aspectratio=169,ignorenonframetext,9pt]{beamer}
\or\documentclass[a4paper,11pt]{article}
   \usepackage{url,beamerarticle}
\or\documentclass[a4paper,11pt]{refart}
   \let\example\relax
   \usepackage{url,beamerarticle}
\fi

\ifcase 0  % choose a theme like these
    % \usetheme{boxes}
    \usetheme{Boadilla}
    % \usetheme{Goettingen}% I recommend
    % \usetheme{Singapore}
    % \usetheme{Pittsburgh}
    % \usetheme{Madrid}
    % \usetheme{Warsaw} % common choice, but often poor
\fi

\usepackage{algorithm,algorithmicx}
\usepackage{graphicx,pgfplots,parskip}
\usepackage{amsmath,amsfonts,amssymb,amsthm,epsfig,epstopdf,url,array}
\usepackage{cite}

\theoremstyle{plain}
\newtheorem{thm}{Theorem}[section]
\newtheorem{lem}[thm]{Lemma}
\newtheorem{prop}[thm]{Proposition}

\theoremstyle{definition}
\newtheorem{defn}{Definition}[section]
\newtheorem{conj}{Conjecture}[section]
\newtheorem{exmp}{Example}[section]
\newtheorem{algo}{Algorithm}[section]

\title{BSTERGMs: Bayesian Separable-Temporal Exponential family Graph Models}
\author{Choi, Seokjun}
\date{14 Dec. 2020}

\begin{document}

\begin{frame}
\maketitle
\end{frame}

\begin{abstract}
Separable Temporal Exponential Random Graph Models (STERGMs) are useful models to describe changes in dynamic networks' structures evolving over time. 
The Bayesian-inference based approach for the model, however, had not been developed yet. 
In this paper, I propose the Bayesian STERGMs (BSTERGMs): STERGMs with the Bayesian setting, and I illustrate the computational algorithm that generates posterior samples from the model dealing with the doubly-intractable constant problem. 
Furthermore, I show the interpretability of the model in the Bayesian context.

\end{abstract}

% \begin{frame}{Outline}
% \tableofcontents
% \end{frame}

\section{Introduction}
Nowadays, new kinds of data having special forms emerge in many field.
These new data are not confined by a series of values of real numbers as traditional data are in the past. 
The data may have non-number type of value, special structure, or even integrated different type of values with model-specific relation.
These data are called complex data, and an attention of statistical analysis on the data increases over time.

As one form of the complex data, networks get highlighted attention among complex data for statistical analysis.
Many phenomena can be represented by some relational structure, and the structure can be expressed effectively
by a form of a network. For example, social networks including both offline and online can be thought as a sort of networks.
In this view, some social aspects of relation between actors can be modeled with an object of social science
including an analysis of power structure and a summary of intimacy structure on given group.
Not only that, application to a transmitted disease leads a modeling using real data in epidemiology.

Since statistical models on network provide not only lots of possibility of application but also
interesting and useful features on observed network, statistical models have important role.
The statistical model gives information about local relations like formation or dissolution between
two actors as well as global features of the whole network such as probability that a specific structure
realizes. Not only that, the statistical models investigate a complex behavior of network including
how the global features of an observed network may be related to local network structure.
During the investigation, statistical model can capture existing dependency on a given network
and provide a tool to understand what certain structural features are more common on rare.

Exponential-family random graph models(ERGMs) may be the most useful tools to analyze cross-sectional networks and their dependencies.
\cite{RN117}, \cite{RN123}, \cite{RN124} and \cite{RN122} proposed the models and apply them to analyze dependent structures,
and the Bayesian approaches of the models were started by \cite{RN115} and \cite{RN116}.
In the ERGMs, The dependencies are captured using a series of local features called network statistics from classical graph-theory.
Since these models play important role in this paper, I will introduce more precisely.

Next natural question is a way to model a network sequence, or the evolution of network over time.
For instance, one may want to model social network structures for 1 year, or
simulation of a disease infection based on the sequential observations in the past.
The first exploration applying ERGMs was \cite{RN109}. 
After them, a notable contributions were of \cite{RN125} and \cite{RN126}, the Temporal ERGMs(TERGMs).
In the TERGM, ERGMs are used by modelling the transition probability between times.
Moreover, \cite{RN93} extend the TERGMs to the Separable Temporal ERGMs(STERGMs), considering an
separation between duration - how long ties tend to last - and incidence - the rate at which new ties are formed -
from prevalence - for example, the total number of ties - of the network properties.
Imposing a conditional independence condition between the duration and incidence,
the models can be extended effectively using similar fitting methods with TERGMs
without adding much complexity.

Although the STERGMs give us strength when it comes to interpret the fitted parameter,
a Bayesian approach for the STERGMs has not been conducted. It is partly because of a hardship
coming from the huge computation in a parameter estimation method for the model.
Nonetheless, the Bayesian approach of STERGMs provides the statistical inference method
following the standard Bayesian way, so the approach has huge advantage of convenience
for statistical works for fitted parameter and predictions to next time network structure.
Thus, in this paper, I specify the bayesian separable temporal exponential random graph models(BSTERGMs)
and propose a fitting algorithm for the model parameter. 
I also discuss ways of inference, the diagnostics procedure, and checking the goodness of fit.

In chapter 2, I review the background meterials including ERGMs, BERGMs, TERGMs, and STERGMs,
and in chapter 3, I show the detail setting of BSTERGMs. In chapter 4, I propose the fitting method for BSTERGM.
In chapter 5, I explain ways to conduct diagnostics, statistical inference, and goodness of fit check.
In chapter 6, I illustrate an application of BSTERGMs to network data of scholars and faculty
from the UCLA medical working project team. Finally, I discuss some extension and 
ways to supplement the BSTERGM further for future research.

\section{Background}
\subsection{Random graphs}
For statistical analysis on relations between subjects of data,
I will introduce terminologies and notations for denoting components related to networks, 
which are a representation of these type of data.

We say each subject on the data to a node. They are same as a vertex in graph theory.
In addition, we say a tie joining two nodes to an edge. A tie can be directed or not.
These two components have fundamental role describing the data.
Define a graph or a network to $y=(V,E)$, where $V$ is a set of node and $E$ is a set of edges of a network.

For simplicity, I will deal with a situation without weight on each node or edge.
Then, a graph $y$ can be represented by simpler form.
Assume that the number of node $n\in \mathbb{N}$ on data is given.
Then, for $\mathcal{B}=\{0,1\}$,
the set $\mathcal{Y} \subset \mathcal{B}^{n^2}$ is a set of graphs of $n$ nodes.
Now, a graph $y \in \mathcal{Y}$ can be represented by a $n \times n$ binary matrix
with elements $y_{ij}=1$ if there is a tie from $i$-th node to $j$-th node,
otherwise $y_{ij}=0$.

If edges of $y$ have directions, then $y$ is called a directed graph. Otherwise, $y$ is called a undirected graph.
These are obvious that, for given $n\in \mathbb{N}$,
$|\mathcal{Y}|=2^{n(n-1)}$ if $\mathcal{Y}$ is the set of all directed graphs not permitting self-connecting edges,
and $|\mathcal{Y}|=2^{n(n-1)/2}$ if $\mathcal{Y}$ is the set of all graphs of undirected one.
Thus, the size of $\mathcal{Y}$ grows exponentially when n increases.

I extend the concept of graphs to be able to describe a situation that connections among nodes are random.
Let $\Omega$ be an event set of all possible connected situations among given nodes.
We say $Y: \Omega \to \mathcal{Y}$ is a random variable for a graph, or a random graph.
For a random graph $Y \in \mathcal{Y}$, denote the edge between i-th node and j-th node by $Y_{ij}$ for $i,j=1,2,...,n$
satisfying $Y_{ij}=1$ if the edge is connected. Otherwise, $Y_{ij}=0$.
Note that the $Y_{ij}$s are also random variables.

@@

\subsection{ERGMs: Exponential random graph models}
From the definition of random graph, one of the natural questions is 
how we can model a distribution on $\mathcal{Y}$.
As a answer, the Exponential random graph models (ERGMs) (\cite{RN123} and \cite{RN122})
gives a classical but 
flexible scheme to model for time-invariant networks represented as a single directed or undirected graph.
Moreover, ERGMs provide a statistical approach to the systematic exploration of several local features simultaneously and 
how these local structures interact to form a global network.

Specifically, following the notation of previous section,
let $\mathcal{Y}$ be a set of random graphs with $n$ nodes and $Y \in \mathcal{Y}$ be a random graph.
Set a distribution on $\mathcal{Y}$ to
    \[P(Y=y;\theta) = \frac{exp(\theta^{T}s(y))}{c(\theta)}\]
for some $\theta\in\mathbb{R}^p$,
where $s(y)\in\mathbb{R}^p$ is a vector which is part of y's sufficient network statistics,
and $c(\theta)\in\mathbb{R}$ is a normalizing constant satisfying $c(\theta)=\sum_{y\in\mathcal{Y}}exp(\theta^{T}s(y))$.
Models which have such a form called the ERGMs.

Note that the form of the right-hand side is the canonical form of exponential family distributions.
That is why the model is called exponential. Moreover, observe that the normalizing constant cannot be directly calculated
because the size of $\mathcal{Y}$ grows exponentially as $n$ grows.
Thus in real data analysis, the calculation of $c(\theta)$ should be avoided.
For more detail of fitting algorithm, see \cite{RN104} the detail on fitting algorithm using MCMCMLE.
You can also see \cite{RN100} addressing the ergm package in R with the MCMCMLE or the classic method using psudolikelihood.

\subsection{BERGMs: Bayesian ERGMs}
\cite{RN115} and \cite{RN116} extend ERGMs to bayesian setting. Putting prior $p(\theta)$ to the parameter $\theta$, the model becomes
\[p(\theta|y)=\frac{p(y|\theta)p(\theta)}{p(y)}=\frac{exp(\theta^T s(y))}{c(\theta)}\frac{p(\theta)}{p(y)}\]

Observe that not only $c(\theta)$ but also the $p(y)$ cannot be evaluated directly
because of the difficulty of summation when $n$ grows and of integration of $p(y)$.
Moreover, the constant depends on the parameter $\theta$, so ordinary MCMC technique cannot be used.
This situation is said to doubly-intractable constant problem(\cite{RN119}), and causes serious challenge to estimate the parameter vector $\theta$.
Thus, The estimation of $\theta$ needs more advanced algorithm than the standard MCMC procedure.
The solution of Caimo and Friel is using the exchange algorithm. See \cite{RN115} for details of fitting methods of BERGMs
and \cite{RN127} for the exchange algorithm.

\subsection{TERGMs: Temporal Exponential family Random Graphs Models}
Let's consider a sequence of random graph $\{Y_1, Y_2, ..., Y_T\} (T\in\mathbb{N})$ and assume that
we are interested in the dynamics or evolution of the network sequence.
The temporal Temporal Exponential family Random Graphs Models(TERGMs)
proposed by \cite{RN125} and \cite{RN126}
is a model for our interest if we permit a Markov assumption
\[P(Y_2,Y_3,...,Y_T|Y_1)=P(Y_2|Y_1)P(Y_3|Y_2)...P(Y_T|Y_{T-1})\]
and focus on the transition probability along index of the sequence.
For convenience, I refer the index as a time.

Specifically, let $\mathcal{Y}$ be a set of graphs with $n$ nodes. 
Let $Y_1=y_1 \in \mathcal{Y}$ be given and $Y_2,...,Y_T (T\in\mathbb{N})$ be random graphs in $\mathcal{Y}$.
Set a distribution on $\mathcal{Y}\times ... \times \mathcal{Y}$ ($T-1$ folds) to
\[P(Y_t=y_t|Y_{t-1}=y_{t-1};\theta) = \frac{exp(\theta^{T}s(y_t, y_{t-1}))}{c(\theta, y_{t-1})}\]
where $t=2,...,T$, $\theta\in\mathbb{R}^p$,
sufficient network statistics $s(y_t, y_{t-1})\in\mathbb{R}^n$,
and a normalizing constant $c(\theta, y_{t-1})=\sum_{y\in\mathcal{Y}}exp(\theta^{T}s(y, y_{t-1}))$.
Models which have a such form called the TERGMs.

The estimation of $\theta$ can be performed by MCMCMLE,
but the estimation suffers severe degeneracy problem.
Furthermore, When we interpret the estimated TERGMs parameters,
we cannot separate the generation and duration because
the higher value of $\theta$ increases both.
These raise demand for modified models to remedy the degeneracy and enable separable interpretation.

\subsection{STERGMs: Separable-Temporal Exponential family Random Graphs Models}
One solution to remedy the problems of TERGMs is 
the Separable-Temporal Exponential family Random Graphs Models, or simply STERGMs. %ref
The key idea is to separate the formation models and dissolution model
for decoupling an effect on edge's duration of model parameter from edge's incidence
when setting a distribution on $\mathcal{Y}\times ... \times \mathcal{Y}$ ($T-1$ folds, $T\in\mathbb{N}$),
where $\mathcal{Y}$ is a set of graphs with given $n$ nodes.

Precisely,
Let $Y_1=y_1 \in \mathcal{Y}$ be given and $Y_2,...,Y_T$ be random graphs (on $\mathcal{Y}$).
Let $\mathcal{Y}^+|_t$ be a subset of $\mathcal{Y}$ consisting all graphs which have equal or additional edges comparing to $y_{t-1}$.
Likewise, let $\mathcal{Y}^-|_t$ be a subset of $\mathcal{Y}$ consisting all graphs which have equal or sparse edges comparing to $y_{t-1}$.
Next, for $Y_t^+: \Omega \to\mathcal{Y}^+|_t$, $Y_t^-: \Omega \to\mathcal{Y}^-|_t$, $y_t^+ \in \mathcal{Y}^+|_t$ and $y_t^- \in \mathcal{Y}^-|_t$, set
\[P(Y_t^+=y_t^+|Y_{t-1}=y_{t-1};\theta^+) = \frac{exp((\theta^+)^{T}s(y_t^+, y_{t-1}))}{c(\theta^+, y_{t-1})}\]
\[P(Y_t^-=y_t^-|Y_{t-1}=y_{t-1};\theta^-) = \frac{exp((\theta^-)^{T}s(y_t^-, y_{t-1}))}{c(\theta^-, y_{t-1})}\]
for some $\theta^+,\theta^-\in\mathbb{R}^p$, $s(y_t^+, y_{t-1}), s(y_t^-, y_{t-1})\in\mathbb{R}^n$, which are parts of sufficient network statistics,
and normalizers $c(\theta^+, y_{t-1})=\sum_{y^+\in\mathcal{Y}^+}exp((\theta^+)^{T}s(y^+, y_{t-1}))$, 
and $c(\theta^-, y_{t-1})=\sum_{y^-\in\mathcal{Y}^-}exp((\theta^-)^{T}s(y^-, y_{t-1}))$.

@@

Additionally, assume the first-order Markov assumption
\[P(Y_2,...,Y_T|Y_1)=P(Y_2|Y_1)...P(Y_T|Y_{T-1})\]
and the conditional independence between $Y_t^+$ and $Y_t^-$ for all $t=2,...,T$, called the separability.
Then we get
\[P(Y_t=y_t|Y_{t-1}=y_{t-1};\theta^+,\theta^-)=P(Y_t^+=y_t^+|Y_{t-1}=y_{t-1};\theta^+) P(Y_t^-=y_t^-|Y_{t-1}=y_{t-1};\theta^-)\]
Models which have a such form called the STERGMs.

\subsection{Network Statistics for ERGMs and derived models}
Now, I briefly discuss the terms of network statistics.
As a usual, $s(y)$ of ERGMs or $s(y_{t}^.,y_{t-1})$ of TERGMs and STERGMs would 
reflect the structure of given network without identifying each node index.
Common candidates of the statistics are here.

>>>>

\documentclass[a4paper, 11pt]{report}
\usepackage{times,epsfig,calc,subfigure}
\usepackage[authoryear,round]{natbib}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{multicol}
%\usepackage[tablesfirst, nolists]{endfloat}
\usepackage{amsbsy, epsfig, epsf, psfrag, booktabs, graphicx, amssymb, enumerate}
\usepackage[hidelinks,bookmarks=false]{hyperref}
\usepackage{lscape}
\usepackage{bm}
\usepackage{float}
\floatplacement{table}{!htbp}
\usepackage{adjustbox}
\usepackage{color, colortbl}

\definecolor{Gray}{gray}{0.9}

\usepackage{makecell}
\usepackage{changepage}%adjust
\usepackage{setspace}%stretch
\usepackage{booktabs} 
\usepackage{multirow}
%\usepackage{hypernat}
\usepackage{authblk}
\usepackage{rotating}
\usepackage{multicol}
\usepackage{epstopdf}
\usepackage{tabularx, caption}
\usepackage[flushleft]{threeparttable}
\usepackage{grad}

\usepackage{kotex}
\usepackage{placeins}

\theoremstyle{definition}
\newtheorem{algo}{Algorithm}[section]

\makeatletter
\newcommand{\distas}[1]{\mathbin{\overset{#1}{\kern\z@\sim}}}%
\newsavebox{\mybox}\newsavebox{\mysim}
\newcommand{\distras}[1]{%
	\savebox{\mybox}{\hbox{\kern3pt$\scriptstyle#1$\kern3pt}}%
	\savebox{\mysim}{\hbox{$\sim$}}%
	\mathbin{\overset{#1}{\kern\z@\resizebox{\wd\mybox}{\ht\mysim}{$\sim$}}}%
}
\makeatother

\renewcommand{\baselinestretch}{1}
\linespread{1.4}
\renewcommand{\arraystretch}{1.3}
\newcommand{\cov}{\mathop{\rm cov}\nolimits} %covariance
\newtheorem{thm}{Theorem}[section]
\newtheorem{prop}[thm]{Proposition}
\newtheorem{defn}[thm]{Definition}
\newcommand\defeq{\mathrel{\overset{\makebox[0pt]{\mbox{\normalfont\tiny\sffamily def}}}{=}}}
\DeclareMathOperator*{\argmin}{argmin}

%%%%%%%%%%%%%%%%%%%%%%%
%   TITLE
%%%%%%%%%%%%%%%%%%%%%%%

\title{BSTERGMs: Bayesian Separable-Temporal Exponential family Graph Models}
\author{Seokjun Choi}
\date{December 2020}

\begin{document}	
\maketitle \pagestyle{plain}

\newpage
\thispagestyle{empty}

  \null
  \begin{center}
    \vskip -1.4cm
    {\fontsize{14pt}{14pt}\selectfont This certifies that Master's thesis of Seokjun Choi is approved.}
    \vskip 4cm%
    {\hskip 5cm\fontsize{12pt}{12pt}\selectfont \underline{\hskip 8cm}}\vspace*{0.2cm}\\
    \hskip 5cm\fontsize{12pt}{12pt}\selectfont Thesis Supervisor: Prof. Ick Hoon Jin\\
    \vskip 2cm
    {\hskip 5cm\fontsize{12pt}{12pt}\selectfont \underline{\hskip 8cm}}\vspace*{0.2cm}\\
    \hskip 5cm\fontsize{12pt}{12pt}\selectfont Committee Member: Prof. Chul Eung Kim\\
    \vskip 2cm
    {\hskip 5cm\fontsize{12pt}{12pt}\selectfont \underline{\hskip 8cm}}\vspace*{0.2cm}\\
    \hskip 5cm\fontsize{12pt}{12pt}\selectfont Committee Member: Prof. Jaewoo Park\\
    \vskip 3cm%
    {\fontsize{14pt}{14pt}\selectfont {The Graduate School \vskip 1.0em Yonsei University \vskip 1.0em Dec 2020}}%
  \end{center}
  \par
\clearpage

\pagestyle{plain}

\clearpage \pagenumbering{roman} \tableofcontents \clearpage

\clearpage \addcontentsline{toc}{chapter}{\protect\numberline{}
	\hspace*{-0.29in} List of Figures}
\listoffigures

\clearpage \addcontentsline{toc}{chapter}{\protect\numberline{}
    \hspace*{-0.29in} List of Tables}
\listoftables

\clearpage \addcontentsline{toc}{chapter}{\protect\numberline{}
    \hspace*{-0.29in} Abstracts}

%%%%%%%%%%%%
% ABSTRACT %
%%%%%%%%%%%%

\begin{abstract}
\begin{center}
{\LARGE 
BSTERGMs: The Bayesian Separable-Temporal Exponential family Graph Models
}
\end{center}
\indent
\vskip 0.8cm
\noindent
\vskip 0.2cm

The separable Temporal Exponential Random Graph Models (STERGMs)(\cite{RN93}) are useful models to describe changes in dynamic networks' structures evolving over time. 
The Bayesian-inference based approach for the models, however, had not been developed yet. 
In this paper, I propose the Bayesian Separable Temporal Exponential Random Graph Models (BSTERGMs): STERGMs with the Bayesian setting, 
and illustrate a computational algorithm that generates posterior samples from the models dealing with the doubly-intractable constant problem. 
Furthermore, I show an inference method of the models in the Bayesian context.

\vspace{\stretch{1}} \noindent
\hrulefill\\
{\bf Key words}
\parbox[t]{0.8\textwidth}
{ : Networks, Exponential Random Graph Models, Bayesian inference, Marcov chain Monte Carlo, Longitudinal}
\end{abstract}

%--------------------------------------------------------------------------------%
%--------------------------------------------------------------------------------%
\clearpage
\chapter{Introduction} \label{Intro}
\pagenumbering{arabic}

Nowadays, new kinds of data having special forms emerge in many fields.
A series of real number values do not confine these new data as traditional data are in the past.
The data may have a non-number type of value, unique structure, or even integrated different types of values with model-specific relation.
These data are called complex data, and the attention of statistical analysis on the data increases over time.

As one form of complex data, networks, or random graphs get highlighted attention among complex data for statistical analysis.
A relational structure can represent many phenomena in the real world, and the structure can be expressed effectively by a form of a network. 
For example, social networks, including offline and online, can be thought of as networks.
In this view, some social aspects of the relationship between actors can be modeled under an object of social science, 
including an analysis of power structure and a summary of intimacy structure on a given group.
Not only that, application to a transmitted disease in epidemiology enables statistical modeling based on real data.

The ability to capture distinct features on the observed network enlarges statistical models' usefulness on networks. 
The statistical models give information about local relations like formation or dissolution between two actors 
as well as global features of the whole network, such as the probability that a specific structure realizes. 
The statistical models investigate the complex behavior of networks, 
including how an observed network's global features may be related to local network structure. 
During the investigation, the statistical models can capture existing dependency on a given network and provide a tool 
to understand what certain structural features are more common on rare.

Exponential-family random graph models(ERGMs) may be the most useful tools to analyze cross-sectional networks and their dependencies.
\cite{RN117}, \cite{RN123}, \cite{RN124} and \cite{RN122} proposed the models and apply them to analyze dependent structures, 
and \cite{RN115} and \cite{RN116} started the Bayesian approaches of the models.
In the ERGMs, The dependencies are captured using a series of local features called network statistics from classical graph-theory. 
Since these models play an important role throughout this paper, I will introduce them more precisely in the following chapter.

The next natural question is a way to model a network sequence or network evolution over time. 
For instance, one may want to model social network structures for one year or 
simulate a disease infection based on the sequential observations in the past. 
The first exploration applying ERGMs to a network sequence was \cite{RN109}. 
After them, a notable contributions were of \cite{RN125} and \cite{RN126}, the Temporal ERGMs(TERGMs). 
In the TERGMs, ERGMs are used by modeling the transition probability between times. 
Moreover, \cite{RN93} extend the TERGMs to the Separable Temporal ERGMs(STERGMs), 
considering the separation between duration - how long ties tend to last - and 
incidence - the rate at which new ties are formed -from 
prevalence - for example, the total number of ties - of the network properties. 
By imposing a conditional independence condition between the duration and incidence on the model, 
the models can be extended to more effectively interpretable ones using similar fitting methods with TERGMs 
without adding much complexity.

Although the STERGMs give us strength when interpreting the fitted parameter, 
a Bayesian approach for the STERGMs has not been conducted. 
It is partly because of the hardship from the massive computation in a parameter estimation method for the models. 
Nonetheless, the Bayesian approach of STERGMs provides the statistical inference method following the standard Bayesian way, 
so the approach has a huge advantage of convenience for statistical works for fitted parameters and predictions to next time network structure. Thus, in this paper, I specify the bayesian separable temporal exponential random graph models(BSTERGMs) and propose a fitting algorithm for the model parameter. I also discuss ways of inference, the diagnostics procedure, and checking the goodness of fit.

Chapter 2 reviews the background materials, including the ERGMs, the BERGMs, the TERGMs, and the STERGMs. 
In chapter 3, I propose the BSTERGMs and show the detailed setting of BSTERGMs. 
In the following chapter, chapter 4, I offer the estimation method for the BSTERGMs. 
Chapter 5 explains ways to conduct diagnostics, statistical inference, and goodness of fit checks. 
Besides, I illustrate an application of BSTERGMs to network data of scholars and faculty
from the UCLA medical working project team in chapter 6, 
Finally, I discuss some possible extensions and ways to supplement the BSTERGMs further for future research.

%--------------------------------------------------------------------------------%
%--------------------------------------------------------------------------------%

\chapter{Background} \label{Chapter2}
\section{Random graphs}
Networks or random graphs play a central role as a general model 
describing relations among subjects in the statistical context. 
To begin with, I will briefly summarize some terminologies and notations for concepts related to the networks.

I will use two words, 'graphs' and 'network,' as the same meaning without distinction. 
We say each subject on the data to a node. 
They sometimes are called vertex in graph theory. 
In addition, we say a tie joining two nodes to an edge. A tie can be directed or not. 
These two components have a fundamental role in describing the relational structure. 
Define a graph or a network to $y=(V,E)$, where $V$ is a set of nodes, and $E$ is a set of edges of a network.

For simplicity, I will deal with a situation without weight on each node or edge.
Then, a graph $y$ can be represented by a more straightforward form.
Assume that the number of node $n\in \mathbb{N}$ on data is given.
Then, for $\mathcal{B}=\{0,1\}$,
the set $\mathcal{Y} \subset \mathcal{B}^{n^2}$ is a set of graphs of $n$ nodes.
Now, a graph $y \in \mathcal{Y}$ can be represented by a $n \times n$ binary matrix
with elements $y_{ij}=1$ if there is a tie from $i$-th node to $j$-th node,
otherwise $y_{ij}=0$.

If edges of $y$ have directions, then $y$ is called a directed graph. Otherwise, $y$ is called an undirected graph.
These are obvious that, for given $n\in \mathbb{N}$,
$|\mathcal{Y}|=2^{n(n-1)}$ if $\mathcal{Y}$ is the set of all directed graphs not permitting self-connecting edges,
and $|\mathcal{Y}|=2^{n(n-1)/2}$ if $\mathcal{Y}$ is the set of all graphs of undirected one.
Thus, the size of $\mathcal{Y}$ exponentially grows when n increases.

I extend the concept of graphs to describe a situation that connections among nodes are random.
Let $\Omega$ be an event set of all possible connected situations among given nodes.
We say $Y: \Omega \to \mathcal{Y}$ is a random variable for a graph, or a random graph.
For a random graph $Y \in \mathcal{Y}$, denote the edge between i-th node and j-th node by $Y_{ij}$ for $i,j=1,2,...,n$
satisfying $Y_{ij}=1$ if the edge is connected. Otherwise, $Y_{ij}=0$.
Note that the $Y_{ij}$s are also random variables.

@@

\section{ERGMs: Exponential random graph models}
One of the natural questions from the definition of the random graph is 
how we can model a probability distribution on $\mathcal{Y}$.
As a answer, the Exponential random graph models (ERGMs) (\cite{RN123} and \cite{RN122})
gives a classical but flexible scheme to model for time-invariant networks represented as a single directed or undirected graph.
Moreover, ERGMs provide a statistical approach to the systematic exploration of several local features simultaneously and 
how these local structures interact to form a global network.

Specifically, following the notation of the previous section,
let $\mathcal{Y}$ be a set of random graphs with $n$ nodes and $Y \in \mathcal{Y}$ be a random graph.
Set a distribution on $\mathcal{Y}$ to
    \[P(Y=y;\theta) = \frac{exp(\theta^{T}s(y))}{c(\theta)}\]
for some $\theta\in\mathbb{R}^p$,
where $s(y)\in\mathbb{R}^p$ is a vector which is part of y's sufficient network statistics,
and $c(\theta)\in\mathbb{R}$ is a normalizing constant satisfying $c(\theta)=\sum_{y\in\mathcal{Y}}exp(\theta^{T}s(y))$.
Models that have such a form are called the ERGMs.

Note that the form of the right-hand side is a canonical form of exponential family distributions.
That is why the model is called exponential. 
Although the form is familiar, directly calculating the normalizing constant is not possible in practical data analysis 
because the size of $\mathcal{Y}$ grows exponentially as $n$ grows. 
Thus for the application to the real data, the calculation of $c(\theta)$ should be avoided.
Luckily, some fitting algorithms have been developed for this issue. 
If you need, see \cite{RN100} addressing the ergm package in R with the MCMCMLE or the classic method using pseudolikelihood. 
For more detail of the MCMCMLE algorithm, see \cite{RN104}

\section{BERGMs: Bayesian ERGMs}
\cite{RN115} and \cite{RN116} extend ERGMs to bayesian setting. 
Putting prior $p(\theta)$ to the parameter $\theta$, the model becomes
\[p(\theta|y)=\frac{p(y|\theta)p(\theta)}{p(y)}=\frac{exp(\theta^T s(y))}{c(\theta)}\frac{p(\theta)}{p(y)}\]

Observe that not only $c(\theta)$ but also the $p(y)$ cannot be evaluated directly
because of the difficulty of summation when $n$ grows and of integration of $p(y)$.
Moreover, the constant depends on the parameter $\theta$, so the ordinary MCMC technique cannot be used.
This situation is said to the doubly intractable constant problem(\cite{RN119}), 
and causes a serious challenge to estimate the parameter vector $\theta$.
Thus, The estimation of $\theta$ needs an advanced algorithm than the standard MCMC procedure.
The solution of Caimo and Friel is using the exchange algorithm. 
You can see \cite{RN115} for details of fitting methods of BERGMs
and \cite{RN127} for the exchange algorithm.

\section{TERGMs: Temporal Exponential family Random Graphs Models}
Let's consider a sequence of random graph $\{Y_1, Y_2, ..., Y_T\} (T\in\mathbb{N})$ and assume that
we are interested in the dynamics or evolution of the network sequence.
The temporal Temporal Exponential family Random Graphs Models(TERGMs)
proposed by \cite{RN125} and \cite{RN126}
is a model for our interest if we permit a Markov assumption
\[P(Y_2,Y_3,...,Y_T|Y_1)=P(Y_2|Y_1)P(Y_3|Y_2)...P(Y_T|Y_{T-1})\]
and focus on the transition probability along index of the sequence.
For convenience, I refer to the index as a time.

Specifically, let $\mathcal{Y}$ be a set of graphs with $n$ nodes. 
Let $Y_1=y_1 \in \mathcal{Y}$ be given and $Y_2,...,Y_T (T\in\mathbb{N})$ be random graphs in $\mathcal{Y}$.
Set a distribution on $\mathcal{Y}\times ... \times \mathcal{Y}$ ($T-1$ folds) to
\[P(Y_t=y_t|Y_{t-1}=y_{t-1};\theta) = \frac{exp(\theta^{T}s(y_t, y_{t-1}))}{c(\theta, y_{t-1})}\]
where $t=2,...,T$, $\theta\in\mathbb{R}^p$,
sufficient network statistics $s(y_t, y_{t-1})\in\mathbb{R}^n$,
and a normalizing constant $c(\theta, y_{t-1})=\sum_{y\in\mathcal{Y}}exp(\theta^{T}s(y, y_{t-1}))$.
Models that have such a form are called the TERGMs.

The estimation of $\theta$ can be performed by MCMCMLE.
Nevertheless, the estimation suffers severe degeneracy problem.
Furthermore, When interpreting the estimated TERGMs parameters,
we cannot separate the generation and duration because
the parameter value of $\theta$ affects both.
These raise demand for modified models to remedy the degeneracy and enable separable interpretation.

\section{STERGMs: Separable-Temporal Exponential family Random Graphs Models}
One solution to remedy the problems of TERGMs is 
the Separable-Temporal Exponential family Random Graphs Models, or simply STERGMs. %ref
The key idea is to separate the formation models and dissolution model
for decoupling an effect on edge's duration of the model parameter from edge's incidence
when setting a distribution on $\mathcal{Y}\times ... \times \mathcal{Y}$ ($T-1$ folds, $T\in\mathbb{N}$),
where $\mathcal{Y}$ is a set of graphs with given $n$ nodes.

Precisely,
Let $Y_1=y_1 \in \mathcal{Y}$ be given and $Y_2,...,Y_T$ be random graphs (on $\mathcal{Y}$).
Let $\mathcal{Y}^+|_t$ be a subset of $\mathcal{Y}$ consisting of all graphs which have equal or additional edges comparing to $y_{t-1}$.
Likewise, let $\mathcal{Y}^-|_t$ be a subset of $\mathcal{Y}$ consisting all graphs which have equal or sparse edges comparing to $y_{t-1}$.
Next, for $Y_t^+: \Omega \to\mathcal{Y}^+|_t$, $Y_t^-: \Omega \to\mathcal{Y}^-|_t$, $y_t^+ \in \mathcal{Y}^+|_t$ and $y_t^- \in \mathcal{Y}^-|_t$, set
\[P(Y_t^+=y_t^+|Y_{t-1}=y_{t-1};\theta^+) = \frac{exp((\theta^+)^{T}s(y_t^+, y_{t-1}))}{c(\theta^+, y_{t-1})}\]
\[P(Y_t^-=y_t^-|Y_{t-1}=y_{t-1};\theta^-) = \frac{exp((\theta^-)^{T}s(y_t^-, y_{t-1}))}{c(\theta^-, y_{t-1})}\]
for some $\theta^+,\theta^-\in\mathbb{R}^p$, $s(y_t^+, y_{t-1}), s(y_t^-, y_{t-1})\in\mathbb{R}^n$, which are parts of sufficient network statistics,
and normalizers $c(\theta^+, y_{t-1})=\sum_{y^+\in\mathcal{Y}^+}exp((\theta^+)^{T}s(y^+, y_{t-1}))$, 
and $c(\theta^-, y_{t-1})=\sum_{y^-\in\mathcal{Y}^-}exp((\theta^-)^{T}s(y^-, y_{t-1}))$.

@@

Additionally, assume the first-order Markov assumption,
\[P(Y_2,...,Y_T|Y_1)=P(Y_2|Y_1)...P(Y_T|Y_{T-1})\]
and the conditional independence between $Y_t^+$ and $Y_t^-$ for all $t=2,...,T$, called the separability.
Then we get
\[P(Y_t=y_t|Y_{t-1}=y_{t-1};\theta^+,\theta^-)=P(Y_t^+=y_t^+|Y_{t-1}=y_{t-1};\theta^+) P(Y_t^-=y_t^-|Y_{t-1}=y_{t-1};\theta^-)\]
Models that have such a form are called the STERGMs.

\section{Network Statistics for ERGMs and derived models}
Now, I briefly discuss the terms of network statistics.
As a usual, $s(y)$ of ERGMs or $s(y_{t}^.,y_{t-1})$ of TERGMs and STERGMs would 
reflect the structure of given network without identifying each node index.
Typical candidates of the statistics are here.

Let $y\in\mathcal{Y}$ and $n\in\mathbb{N}$ be the number of nodes of $y$.
The most straightforward candidate is the number of edges, denoted by $|y|$.

Following \cite{RN104}, usual candidates are distributions defined by some sense of the structure of a given graph. 
For a given $i, 1\leq i \leq n-1$, $D_i(y)$ is defined as the number of nodes in y whose the number of edges incident to the node equals $i$, 
and called $i$-th order node degree distribution.

For a given $i, 1\leq i \leq n-2$, $P_i(y)$ is defined to be the number of dyads $(j,k), 1\leq j,k \leq n$
in y whose $j$ and $k$ have nodes connected by an edge each other and they share exactly $i$ connected nodes in common,
and called $i$-th order edgewise shared partner distribution.
